---
title: "Project1"
author: "Sally Sydnor, River Robins, Grace FitzGerald, Kyle Kendall"
date: "2023-10-11"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

# Cleaning, Sourcing, Loading

## Loading Libraries

```{r, include =FALSE}
# tidyverse libraries
library(tinytex)
library(ggplot2)
library(readr)
library(dplyr)
library(stringr)
library(GGally)
library(psych)
library(lattice)
library(ggpubr)
library(data.table)
library(plyr)
library(scales)
library(grid)
library(ggpubr)
library(gplots)
library(lattice)
library(knitr)
library(here)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(ggfortify)
library(MASS)
library(lindia)
library(olsrr)
```

## Sourcing Directories

```{r, include =FALSE}
traindir <- ("~/UVA/Systems/SYS 4021 Linear Statistical Models/R/TrainData")
sourcedir <- ("~/UVA/Systems/SYS 4021 Linear Statistical Models/R")


# set the working directory to traindir
setwd(traindir)
#check the current working directory
#it should be same as your traindir directory
getwd()
```

## Sourcing Data

```{r, include =FALSE}
# Source AccidentInput and PCAplots
source("AccidentInput.R")
source("PCAplots.R")

acts <- file.inputl(traindir)

# Next a data frame with all accidents from all years from 2001 - 2019
#combined data frame
totacts <- combine.data(acts)
```

## Cleaning

```{r, include =FALSE}
# For ACCDMG
dmgbox <- ggplot(totacts, aes(y=ACCDMG)) + geom_boxplot()
#dmgbox

# find only those above the upper whisker
upper <- ggplot_build(dmgbox)$data[[1]]$ymax
xdmg <- totacts[totacts$ACCDMG > upper,]

# For Casualties (TOTINJ + TOTKLD)

xdmg$Casualties <- xdmg$TOTINJ + xdmg$TOTKLD

# Remove 9/11
xdmg <- xdmg[-181,]


```

```{r, include =FALSE}
# Remove duplicates
xdmgnd <- xdmg[!(duplicated(xdmg[, c("INCDTNO", "YEAR", "MONTH", "DAY", "TIMEHR", "TIMEMIN")])),]

#Reset rownames (observation #s) for sequential numbering- otherwise they will remain the #s from totacts

rownames(xdmgnd) <- NULL
```

```{r, include =TRUE, tidy = TRUE}
#categorical variable setup

xdmgnd$TYPE <- factor(xdmgnd$TYPE, labels = c("Derailment", "HeadOn", 
                "Rearend", "Side", "Raking", "BrokenTrain", "Hwy-Rail", 
                  "GradeX", "Obstruction", "Explosive", "Fire","Other",
                  "SeeNarrative" ))

xdmgnd$TYPEQ <- factor(xdmgnd$TYPEQ, labels = c("NA", "Freight", "Passenger", "Commuter", 
                    "Work",  "Single", "CutofCars", "Yard", "Light", "Maint",
                  "MaintOfWay", "Passenger", "Commuter", "ElectricMulti", "ElectricMulti"))

xdmgnd$WEATHER <- factor(xdmgnd$WEATHER, labels = c("Clear", "Cloudy", "Rain", "Fog", "Sleet", "Snow"))

xdmgnd$Cause <- rep(NA, nrow(xdmgnd))

xdmgnd$Cause[which(substr(xdmgnd$CAUSE, 1, 1) == "M")] <- "M"
xdmgnd$Cause[which(substr(xdmgnd$CAUSE, 1, 1) == "T")] <- "T"
xdmgnd$Cause[which(substr(xdmgnd$CAUSE, 1, 1) == "S")] <- "S"
xdmgnd$Cause[which(substr(xdmgnd$CAUSE, 1, 1) == "H")] <- "H"
xdmgnd$Cause[which(substr(xdmgnd$CAUSE, 1, 1) == "E")] <- "E"
```

Create new variable: Casualty

```{r, include =TRUE}
xdmgnd$Casualty <- xdmgnd$TOTKLD + xdmgnd$TOTINJ
table(xdmgnd$Casualty)

## Build a data frame of accidents with at least one casualty
onecas <- xdmgnd %>%
  filter(Casualty > 0)
onecas <- mutate(onecas, Derail = ifelse(TYPE == "Derailment", "Y", "N"))
```

# Generating Hypotheses

Building off of our in-class work on extreme accidents, choose at least
2 well-formed, actionable hypotheses for each severity metric (ACCDMG
and Casualties) that you plan to test. Explain why the hypotheses are
actionable and demonstrate how you arrived at each hypothesis
(visualizations, summary statistics, and any other supporting evidence).
Write out your null and alternative hypotheses. Consider only
qualitative variables or the interaction between a qualitative and
quantitative variable.

## Hypothesis 1 -- ACCDMG

Thoughts: Accident damage depending on train speed and tons (higher
weight and speed = more damage)

Accidents happening earlier might have more damage (we could've fixed
previous problems)

Accident damage depending on train speed and type of accident
(derailment and high speed)

Accident damage depending on weather and train speed Null hypothesis:
Weather and train speed do not affect accident damage Alternative
hypothesis: Accidents that occur during inclement weather (rain, sleet,
or snow) with high speed have higher accident damage

### Correlation Matrix Graph Replication

```{r}
# scatter plot of ACCDMG with ...
pairs.panels(xdmgnd[,c("WEATHER", "VISIBLTY", "TRNSPD", "TYPE", "ACCDMG")])
```

### Further Hypothesis Development with Box Plots

```{r}
#table(xdmgnd$WEATHER)

ggplot(xdmgnd, aes(x=WEATHER, y=ACCDMG))+
  geom_boxplot()

ggplot(xdmgnd, aes(x=WEATHER))+
  geom_bar()

#unique(xdmgnd$METHOD)

table(xdmgnd$MOPERA)
table(xdmgnd$SIGNAL)
class(xdmgnd$MOPERA)


#MOPERA
ggplot(xdmgnd, aes(x=factor(MOPERA), y=ACCDMG))+
  geom_boxplot()
#SIGNAL
ggplot(xdmgnd, aes(x=factor(SIGNAL), y=ACCDMG))+
  geom_boxplot()

#finding accdmg median with respect to SIGNAL
medianacc <- xdmgnd %>%
  group_by(SIGNAL) %>%
  summarize(meddmg = median(ACCDMG))
medianacc

test2 <- xdmgnd %>%
  group_by(METHOD) %>%
  summarize(avgdmg = mean(ACCDMG)) %>%
  arrange(desc(avgdmg))
test2

```

### Null Hypothesis & Alternative Hypothesis

Accident damage depends on train speed and if it was signaled

Null hypothesis: Accidents occurring at high speeds in non-signaled
territories do not result in more damage than those at high speeds in
signaled territories Alternative hypothesis: Accidents occurring at high
speeds in non-signaled territories result in more damage than those at
high speeds in signaled territories

## Hypothesis 2 -- Casualties

### Correlation Matrix Graph Replication

```{r}
# scatter plot of Casualties with ...
pairs.panels(xdmgnd[,c("WEATHER", "VISIBLTY", "TRNSPD", "TYPE", "Casualties")])
```

### Further Hypothesis Development with Box Plots

```{r}

ggplot(xdmgnd, aes(x=WEATHER, y=Casualties))+
  geom_boxplot()

ggplot(xdmgnd, aes(x=WEATHER))+
  geom_bar()
```

### Null Hypothesis & Alternative Hypothesis

There are more casualties during derailments at high speeds

Null hypothesis: Alternative hypothesis:

# ACCDMG Analysis

## Hypothesis 1: Accidents occurring at high speeds in non-signaled territories do not result in more damage than those at high speeds in signaled territories.

## A) The feature and model selection techniques you used to find appropriate models for this problem

Build linear model to test hypothesis

```{r}
# regular model
accdmg.lm1<-lm(ACCDMG~TRNSPD + SIGNAL,data=xdmgnd)
summary(accdmg.lm1)
# interaction model
accdmg.lm2 <- lm(ACCDMG ~ (TRNSPD + SIGNAL)^2, data = xdmgnd)
summary(accdmg.lm2)
# complete interaction model
accdmg.lm3 <- lm(ACCDMG~(TRNSPD+SIGNAL)^2+I(TRNSPD^2)+I(SIGNAL^2),data=xdmgnd)
summary(accdmg.lm3)
# step model
accdmg.lm4 <- step(accdmg.lm3)
summary(accdmg.lm4)
```

## (b) Your treatment of ordinal and categorical variables (i.e., how were they coded)

```{r}
# TRNSPD: interval variables
# SIGNAL: binary variable
```

## (c) How you assessed your models (e.g., adjusted R2, AIC, etc.)

```{r}
AIC(accdmg.lm1)
AIC(accdmg.lm2)
AIC(accdmg.lm3)
AIC(accdmg.lm4)

summary(accdmg.lm1)$adj.r.squared
summary(accdmg.lm2)$adj.r.squared
summary(accdmg.lm3)$adj.r.squared
summary(accdmg.lm4)$adj.r.squared

# models 3 and 4 have the smallest AIC value
# models 3 and 4 have the largest adjusted R squared values

```

## (d) How you diagnosed problems with the models

```{r}
# cook's distance
autoplot(accdmg.lm1, which=4, ncol = 1, label.size = 3) + theme_bw() #Cook's distance
autoplot(accdmg.lm2, which=4, ncol = 1, label.size = 3) + theme_bw() #Cook's distance
autoplot(accdmg.lm3, which=4, ncol = 1, label.size = 3) + theme_bw() #Cook's distance
autoplot(accdmg.lm4, which=4, ncol = 1, label.size = 3) + theme_bw() #Cook's distance

# point 5251 is above 0.5 on cook's distance plot for models 1 and 2
# point 5251 is above 3 on cook's distance plot for models 3 and 4
# to fix model, point 5251 should be removed

```

```{r}
# residuals vs fitted
autoplot(accdmg.lm1,which=1, ncol = 1, label.size = 3) + theme_bw()#Residual vs. Fitted
autoplot(accdmg.lm2,which=1, ncol = 1, label.size = 3) + theme_bw()#Residual vs. Fitted
autoplot(accdmg.lm3,which=1, ncol = 1, label.size = 3) + theme_bw()#Residual vs. Fitted
autoplot(accdmg.lm4,which=1, ncol = 1, label.size = 3) + theme_bw()#Residual vs. Fitted

# plots for 1 and 2 are similar
# plots for 3 and 4 are same

# testing for constant variance
# Are the residuals homoscedastic
ols_test_breusch_pagan(accdmg.lm1)
ols_test_breusch_pagan(accdmg.lm2)
ols_test_breusch_pagan(accdmg.lm3)
ols_test_breusch_pagan(accdmg.lm4)

# all models have a p value of 0.000, therefore we reject constant variance (assumption of constant variance is violated)
```

```{r}
# qq plot
autoplot(accdmg.lm1,which=2, ncol = 1, label.size = 3) + theme_bw() #QQ
autoplot(accdmg.lm2,which=2, ncol = 1, label.size = 3) + theme_bw() #QQ
autoplot(accdmg.lm3,which=2, ncol = 1, label.size = 3) + theme_bw() #QQ
autoplot(accdmg.lm4,which=2, ncol = 1, label.size = 3) + theme_bw() #QQ

# upper tail is non-gaussian for all models

```

part D cont.

```{r}
#scale location
autoplot(accdmg.lm1,which=3, ncol = 1, label.size = 3) + theme_bw() #Scale-Location
autoplot(accdmg.lm2,which=3, ncol = 1, label.size = 3) + theme_bw() #Scale-Location
autoplot(accdmg.lm3,which=3, ncol = 1, label.size = 3) + theme_bw() #Scale-Location
autoplot(accdmg.lm4,which=3, ncol = 1, label.size = 3) + theme_bw() #Scale-Location

# all plots follow same shape
```

part D cont.

```{r}
# residuals vs leverage
autoplot(accdmg.lm1,which=5,ncol = 1, label.size = 3) + theme_bw() #Redisuals vs. Leverage
autoplot(accdmg.lm2,which=5,ncol = 1, label.size = 3) + theme_bw() #Redisuals vs. Leverage
autoplot(accdmg.lm3,which=5,ncol = 1, label.size = 3) + theme_bw() #Redisuals vs. Leverage
autoplot(accdmg.lm4,which=5,ncol = 1, label.size = 3) + theme_bw() #Redisuals vs. Leverage
```

part D cont.

```{r}
# Cook's distance vs leverage
autoplot(accdmg.lm1,which=6, ncol = 1, label.size = 3) + theme_bw() #Cook's dist vs. Leverage
autoplot(accdmg.lm2,which=6, ncol = 1, label.size = 3) + theme_bw() #Cook's dist vs. Leverage
autoplot(accdmg.lm3,which=6, ncol = 1, label.size = 3) + theme_bw() #Cook's dist vs. Leverage
autoplot(accdmg.lm4,which=6, ncol = 1, label.size = 3) + theme_bw() #Cook's dist vs. Leverage
```

## (e) How you adjusted the models based on these assessment

adjust models

```{r}
# problems found: 
#point with high value cook's distance (remove that point)
xdmgnd <- xdmgnd[-5251,]
```

```{r}
# non-constant variance / gaussian errors (transform the response)
boxcox(accdmg.lm1) # brackets neither 0 or 1, so use optimal lambda

##find max likelihood
max(boxcox(accdmg.lm1, plotit = F)$y)
## find best lambda value
boxcox(accdmg.lm1, plotit = F)$x[which.max(boxcox(accdmg.lm1, plotit = F)$y)] 
##The best lambda and store in L
L<-boxcox(accdmg.lm1, plotit = F)$x[which.max(boxcox(accdmg.lm1, plotit = F)$y)] 
L
##The model with the best lambda transformation
accdmg.lm1.boxcox<-lm((ACCDMG^L-1)/L~TRNSPD+SIGNAL,data=xdmgnd)

##Display regression results for boxcox model
summary(accdmg.lm1)
summary(accdmg.lm1.boxcox)

```

part E cont.

```{r}
# non-constant variance / gaussian errors (transform the response)
boxcox(accdmg.lm3) # brackets neither 0 or 1, so use optimal lambda

##find max likelihood
max(boxcox(accdmg.lm3, plotit = F)$y)
## find best lambda value
boxcox(accdmg.lm3, plotit = F)$x[which.max(boxcox(accdmg.lm3, plotit = F)$y)] 
##The best lambda and store in L
L<-boxcox(accdmg.lm3, plotit = F)$x[which.max(boxcox(accdmg.lm3, plotit = F)$y)] 
L
##The model with the best lambda transformation
accdmg.lm3.boxcox<-lm((ACCDMG^L-1)/L~TRNSPD+SIGNAL,data=xdmgnd)

##Display regression results for boxcox model
summary(accdmg.lm3)
summary(accdmg.lm3.boxcox)
```

## Hypothesis 2: Dusk presents the greatest damage for ACCDMG when controlling for other factors (e.g. TRNSPD)

## A) The feature and model selection techniques you used to find appropriate models for this problem

```{r}

```

## (b) Your treatment of ordinal and categorical variables (i.e., how were they coded)

```{r}

```

## (c) How you assessed your models

## (d) How you diagnosed problems with the models

## (e) How you adjusted models

# Casualties Analysis

## Hypothesis 1: There are more casualties with trains with more passenger loads that derail

## (a) The feature and model selection techniques you used to find appropriate models for this problem

```{r}
onecas %>% ggplot(aes(Casualty)) +
  geom_boxplot()

table(onecas$Derail)

pairs.panels(dplyr::select(onecas, c('Casualty','HIGHSPD','CDTRMIN','LOADP1','LOADP2','TONS','CARS')))

cas.lm <- lm(Casualty ~ LOADP1 + Derail + Derail*LOADP1, data = onecas)
summary(cas.lm)

plot(cas.lm)
```

## (b) Your treatment of ordinal and categorical variables (i.e., how were they coded)

```{r}
## Build a data frame of accidents with at least one casualty
onecas <- xdmgnd %>%
  filter(Casualty > 0)
onecas <- mutate(onecas, Derail = ifelse(TYPE == "Derailment", "Y", "N"))
```

## (c) How you assessed your models (e.g., adjusted R2, AIC, etc.)

```{r}
AIC(cas.lm)
```

## (d) How you diagnosed problems with the models; and

```{r}
onecas_trimmed <- onecas[-c(384,210,2772,2222),]

cas.lm.trimmed <- lm(Casualty ~ LOADP1 + Derail + Derail*LOADP1, data = onecas_trimmed)
summary(cas.lm.trimmed)

plot(cas.lm.trimmed)

plot(cooks.distance(cas.lm.trimmed))

bxcx <- boxcox(cas.lm.trimmed)

bxcx$x[which.max(bxcx$y)]

cas.lm.transformed <- lm((Casualty^(-1.47)-1)/(-1.47) ~ LOADP1 + Derail + Derail*LOADP1, data = onecas_trimmed)
summary(cas.lm.transformed)

plot(cas.lm.transformed)
```

## post transform (c)

```{r}
AIC(cas.lm.trimmed)
AIC(cas.lm.transformed)
```

## (e) How you adjusted the models based on these assessment

```{r}
cas.lm.transformed <- lm((Casualty^(-1.47)-1)/(-1.47) ~ LOADP1 + Derail + Derail*LOADP1, data = onecas_trimmed)
summary(cas.lm.transformed)

plot(cas.lm.transformed)
```

Results: the interaction between derailment and loads did have a
significant effect on casualties

## Hypothesis 2: There are more casualties during derailments at high speeds

## (a) The feature and model selection techniques you used to find appropriate models for this problem

Linear Model

```{r}
#Models
model1<-lm(Casualty~Derail,data=xdmgnd_derail)
model2 <- lm(Casualty~Derail+HIGHSPD, data = xdmgnd_derail)
model3 <-lm(Casualty~(HIGHSPD+Derail)^2,data=xdmgnd_derail)
summary(model1)
```

## (b) Your treatment of ordinal and categorical variables (i.e., how were they coded)

```{r}
#Dummy Variable
xdmgnd_derail <- mutate(xdmgnd, Derail = ifelse(TYPE == "Derailment", "Y", "N"))
```

## (c) How you assessed your models (e.g., adjusted R2, AIC, etc.)

```{r}
#Adjusted R^2
summary(model1)$adj.r.squared
summary(model2)$adj.r.squared
summary(model3)$adj.r.squared
#summary(model4)$adj.r.squared

##AIC:
AIC(model1)
AIC(model2)
AIC(model3)
#AIC(model4)

##BIC:
BIC(model1)
BIC(model2)
BIC(model3)
```

## (d) How you diagnosed problems with the models; and

### Model 1

```{r}
##Generate diagnostics plot one by one
#Plot graphs individually

autoplot(model1,which=1, ncol = 1, label.size = 3) + theme_bw()#Residual vs. Fitted

autoplot(model1,which=2, ncol = 1, label.size = 3) + theme_bw() #QQ

autoplot(model1,which=3, ncol = 1, label.size = 3) + theme_bw() #Scale-Location

autoplot(model1, which=4, ncol = 1, label.size = 3) + theme_bw() #Cook's distance

autoplot(model1,which=5,ncol = 1, label.size = 3) + theme_bw() #Redisuals vs. Leverage

autoplot(model1,which=6, ncol = 1, label.size = 3) + theme_bw() #Cook's dist vs. Leverage

# Are the residuals homoscedastic
ols_test_breusch_pagan(model1)
```

### Model 2

```{r}
##Generate diagnostics plot one by one
#Plot graphs individually

autoplot(model2,which=1, ncol = 1, label.size = 3) + theme_bw()#Residual vs. Fitted

autoplot(model2,which=2, ncol = 1, label.size = 3) + theme_bw() #QQ

autoplot(model2,which=3, ncol = 1, label.size = 3) + theme_bw() #Scale-Location

autoplot(model2, which=4, ncol = 1, label.size = 3) + theme_bw() #Cook's distance

autoplot(model2,which=5,ncol = 1, label.size = 3) + theme_bw() #Redisuals vs. Leverage

autoplot(model2,which=6, ncol = 1, label.size = 3) + theme_bw() #Cook's dist vs. Leverage

# Are the residuals homoscedastic
ols_test_breusch_pagan(model2)
```

### Model 3

```{r}
##Generate diagnostics plot one by one
#Plot graphs individually

autoplot(model3,which=1, ncol = 1, label.size = 3) + theme_bw()#Residual vs. Fitted

autoplot(model3,which=2, ncol = 1, label.size = 3) + theme_bw() #QQ

autoplot(model3,which=3, ncol = 1, label.size = 3) + theme_bw() #Scale-Location

autoplot(model3, which=4, ncol = 1, label.size = 3) + theme_bw() #Cook's distance

autoplot(model3,which=5,ncol = 1, label.size = 3) + theme_bw() #Redisuals vs. Leverage

autoplot(model3,which=6, ncol = 1, label.size = 3) + theme_bw() #Cook's dist vs. Leverage

# # Are the residuals homoscedastic
ols_test_breusch_pagan(model3)
```

## (e) How you adjusted the models based on these assessment

```{r}
#Model 2 is best
model2 <- lm(Casualty~Derail+HIGHSPD, data = xdmgnd_derail)
plot(model2)

```

# Evidence Recommendation to FRA

# References

<https://railroads.dot.gov/divisions/security-preparedness-and-accident-analysis/fra-investigations-railroad-accidents>
The FRA mentions that most accidents assigned for investigation fall
into one of the following categories: rail equipment (collisions and
derailments), highway-rail grade crossing incidents, and fatalities
(employee and contractor). We used derailments as one of our main
variables for making our models about accident damage and casualties.
Since the FRA investigated accidents that had to do with derailments, it
illustrates that derailments have a strong connection with increased
accident damage or casualties.
